Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-07-11 Paper/paper.tex
minor changes

\newcommand{\Caltech}{\affiliation{Theoretical Astrophysics 350-17,
    California Institute of Technology, Pasadena, CA 91125, USA}}
\newcommand{\Cornell}{\affiliation{Center for Radiophysics and Space
    Research, Cornell University, Ithaca, New York 14853, USA}}
\newcommand{\CITA}{\affiliation{Canadian Institute for Theoretical
    Astrophysics, 60 St.~George Street, University of Toronto,
    Toronto, ON M5S 3H8, Canada}} %
\newcommand{\GWPAC}{\affiliation{Gravitational Wave Physics and
    Astronomy Center, California State University Fullerton,
    Fullerton, California 92834, USA}} %
\newcommand{\AEI}{\affiliation{Albert Einstein Institute,
Am M\"uhlenberg, G\"olm, Germany}} %
\newcommand{\CIFAR}{\affiliation{Canadian Institute for Advanced Research, 
    180 Dundas St.~West, Toronto, ON M5G 1Z8, Canada}} %

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}
\caption{We show here probability distributions for mass ratio $\eta$ as measured
for the same injection at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
% 
In each panel: the dashed red line marks the median {\it measured} value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The difference between the true and measured values
is the systematic bias introduced
by ignoring tidal effects in templates. The darker shading shows
the recovered $90\%$ confidence interval for $\eta$, $(\Delta\eta)^{90\%}$, 
which is a direct measure of our statistical uncertainty on $\eta$.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}
% #################
\begin{figure*}
\centering 
\includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNMchirpCIWidths90_0_Lambda_SNR}\\
\includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNEtaCIWidths90_0_Lambda_SNR}\\
\includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNChiBHCIWidths90_0_Lambda_SNR}
\caption{
We show here the statistical uncertainty of our measurement of the 
dimensionless mass ratio $\eta$ (at $90\%$ confidence), over the NSBH
parameter space.
Individual panels show the same as a function of BH mass and spin.
Across each row, we see the effect of increasing signal strength (i.e.
SNR) with the NS's tidal deformability $\lambdans$ fixed. Down each
column, we see the effect of increasing $\lambdans$, at fixed SNR.
}
\label{fig:CIWidths90_Lambda_SNR}
\end{figure*}
% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

\red{SAY HERE THAT WE ONLY USE THE $l=|m|=2$ modes.}
\prayush{
This is possible because of our simplification of 
using only the dominant $l=|m|=2$ modes of the GW strain~\footnote{The 
$l=|m|=2$ multipoles of the GW strain, when decomposed in a spin $-2$ weighted
spherical harmonic basis, contain more than $99\%$ of the total signal power}
that makes various angle parameters completely degenerate with distance. 
}
% 
\subsection{Waveform Models}\label{s2:waveforms}

@@

In this paper we use LEA for our signal and template modeling, but switch the 
underlying BBH model to SEOBNRv2 (and refer to it as enhanced-LEA or
``LEA+'')~\cite{Taracchini:2013rva}. We using the reduced-order
frequency-domain version of SEOBNRv2, which has the additional benefit of
reducing computational cost~\cite{Purrer:2015tud}. We expect this enhancement
from LEA$\rightarrow$LEA+ to make our conclusions more robust because: (a) the 
SEOBNRv2 model is more accurate~\cite{Kumar:2015tha,Kumar:2016dhh}, and (b)
the differences between the two EOB models are caused by the
inaccuracies of SEOBNRv1 during the {\it inspiral} phase, many orbits before 
merger~\cite{Kumar:2015tha}.
Since LEA only augments inspiral phasing with PN tidal terms, our
change in the underlying BBH model does not change LEA's construction, {\it and}
increases the overall model accuracy during inspiral.
% We will refer to this enhanced-LEA model henceforth as ``LEA+''.

@@

\red{SOMEHOW INTRODUCE SNR AND MENTION IF IT IS A NETWORK SNR OR WHAT KIND OF 
'SNR' IS IT?}

The process of measuring systematic and statistical measurement errors
involves simulating many artificial GW signals, and inferring source binary
parameters from them using Bayesian statistics.
We start with generating a signal waveform, using the model LEA+, and injecting
it in zero noise to obtain a stretch of data $d_n$. Source binary parameters
$\vec{\Theta}:=\vec{\theta}\cup\{\lambdans\}$ are reconstructed from this
injected signal. Using Bayes' theorem, their joint inferred probability
distribution of $\vec{\Theta}$ can be written	 as
\begin{equation}\label{eq:postprob}
 p(\vec{\Theta} | d_n, K) = \dfrac{p(d_n|\vec{\Theta}, K)\,p(\vec{\Theta} | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\Theta} | K)$ is the a-priori probability of binary parameters
taking particular values. Throughout this paper, we impose a prior uniform on
individual component masses, BH spin, and NS's tidal deformability. In
addition, we restrict mass-ratios to $q\geq 2$, as LEA+ is not valid for 
$1\leq q\leq 2$. $p(d_n|\vec{\Theta}, K)$ is the
likelihood of obtaining the given stretch of data $d_n$ if we assume that a
signal parameterized by $\vec{\Theta}$ is buried in it, and is given by
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\Theta}) \equiv p(d_n| \vec{\Theta}, K) = \mathcal{N}\, \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\Theta})$ is a filter template with parameters 
$\vec{\Theta}$, and $\langle\cdot|\cdot\rangle$ is a suitably defined
detector-noise weighted inner-product~\footnote{The inner product
$\langle\cdot|\cdot\rangle$  is defined as
\begin{equation}
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
\end{equation}
where $\tilde{a}(f)$ is the Fourier transform of the finite time series $a(t)$,
and $S_n(|f|)$ is the one-sided amplitude spectrum of detector noise.}
The denominator in Eq.~\ref{eq:postprob} is the a-priori probability of finding
the particular signal in $d_n$ and we assume that each injected signal is as
likely as any other. Having constructed the probability distribution function
$p(\vec{\Theta} | d_n, K)$, extracting the measured probability distribution
for a single parameter (say $\alpha$) involves integrating
\begin{equation}\label{eq:marginalize}
 p(\alpha | d_n, K) = \int\D \vec{\Theta}_\alpha\, p(\vec{\Theta} | d_n, K),
\end{equation}
where $\vec{\Theta}_\alpha$ is the set of remaining parameters, i.e.
$\vec{\Theta}_\alpha:=\vec{\Theta} - \{\alpha\}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{How is PE affected if we ignore NS matter effects?}\label{s1:PEwithnoNS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% #################
\begin{figure*}
\centering 
\includegraphics[width=1.6\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
% 
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
}
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\end{figure*}
%
\begin{figure*}[!t]
\centering
\includegraphics[width=2\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR20_70_linear}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
}
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
\begin{figure*}
\centering
\includegraphics[width=1.95\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
}
\label{fig:TN_BHspinBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
Past and future GW search and source characterization efforts with Advanced LIGO
have used (or plan to) BH-BH waveform templates for NS-BH mergers, ignoring
NS tidal effects on the emitted GWs. In this section we present a fully
Bayesian analysis of the effects of this assumption on the recovery of NSBH
masses and spins from their GW signals.

>>>>

\newcommand{\Caltech}{\affiliation{Theoretical Astrophysics 350-17,
    California Institute of Technology, Pasadena, CA 91125, USA}}
\newcommand{\Cornell}{\affiliation{Center for Radiophysics and Space
    Research, Cornell University, Ithaca, New York 14853, USA}}
\newcommand{\CITA}{\affiliation{Canadian Institute for Theoretical
    Astrophysics, 60 St.~George Street, University of Toronto,
    Toronto, ON M5S 3H8, Canada}} %
\newcommand{\GWPAC}{\affiliation{Gravitational Wave Physics and
    Astronomy Center, California State University Fullerton,
    Fullerton, California 92834, USA}} %
\newcommand{\AEI}{\affiliation{Albert Einstein Institute,
Am M\"uhlenberg, Golm, Germany}} %
\newcommand{\CIFAR}{\affiliation{Canadian Institute for Advanced Research, 
    180 Dundas St.~West, Toronto, ON M5G 1Z8, Canada}} %

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION

% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

\subsection{Waveform Models}\label{s2:waveforms}

@@

In this paper we use LEA for our signal and template modeling, but switch the 
underlying BBH model to SEOBNRv2 (and refer to it as enhanced-LEA or
``LEA+'')~\cite{Taracchini:2013rva}. We using the reduced-order
frequency-domain version of SEOBNRv2, which has the additional benefit of
reducing computational cost~\cite{Purrer:2015tud}. We expect this enhancement
from LEA$\rightarrow$LEA+ to make our conclusions more robust because: (a) the 
SEOBNRv2 model is more accurate~\cite{Kumar:2015tha,Kumar:2016dhh}, and (b)
the differences between the two EOB models are caused by the
inaccuracies of SEOBNRv1 during the {\it inspiral} phase, many orbits before 
merger~\cite{Kumar:2015tha}.
Since LEA only augments inspiral phasing with PN tidal terms, our
change in the underlying BBH model does not change LEA's construction, {\it and}
increases the overall model accuracy during inspiral.
% 
Finally, we note that we approximate the full GW signal with its dominant
$l=|m|=2$ modes, that are modeled by LEA+.
% ~\footnote{The 
% $l=|m|=2$ multipoles of the GW strain, when decomposed in a spin $-2$ weighted
% spherical harmonic basis, contain more than $99\%$ of the total signal power}
% This approximation has the added advantage of making various angle parameters
% degenerate with source distance.

@@

The process of measuring systematic and statistical measurement errors
involves simulating many artificial GW signals, and inferring source binary
parameters from them using Bayesian statistics.
We start with generating a signal waveform, using the model LEA+, and injecting
it in zero noise to obtain a stretch of data $d_n$. Source binary parameters
$\vec{\Theta}:=\vec{\theta}\cup\{\lambdans\}$ are reconstructed from this
injected signal. Using Bayes' theorem, their joint inferred probability
distribution of $\vec{\Theta}$ can be written	 as
\begin{equation}\label{eq:postprob}
 p(\vec{\Theta} | d_n, K) = \dfrac{p(d_n|\vec{\Theta}, K)\,p(\vec{\Theta} | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\Theta} | K)$ is the a-priori probability of binary parameters
taking particular values. Throughout this paper, we impose a prior uniform on
individual component masses, BH spin, and NS's tidal deformability. In
addition, we restrict mass-ratios to $q\geq 2$, as LEA+ is not calibrated for 
$1\leq q\leq 2$. $p(d_n|\vec{\Theta}, K)$ is the
likelihood of obtaining the given stretch of data $d_n$ if we assume that a
signal parameterized by $\vec{\Theta}$ is buried in it, and is given by
\begin{equation}\label{eq:likelihood}
 p(d_n| \vec{\Theta}, K) \equiv \LL(\vec{\Theta}) = \mathcal{N}\, \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\Theta})$ is a filter template with parameters 
$\vec{\Theta}$, and $\langle\cdot|\cdot\rangle$ is a suitably defined
detector-noise weighted inner-product~\footnote{The inner product
$\langle\cdot|\cdot\rangle$  is defined as
\begin{equation}
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
\end{equation}
where $\tilde{a}(f)$ is the Fourier transform of the finite time series $a(t)$,
and $S_n(|f|)$ is the one-sided amplitude spectrum of detector noise.}.
\red{[MP: INSERT A LINE ABOUT APPROXIMATING $\LL \approx e^{\rho^2/2}$}
The denominator in Eq.~\ref{eq:postprob} is the a-priori probability of finding
the particular signal in $d_n$ and we assume that each injected signal is as
likely as any other. Having constructed the probability distribution function
$p(\vec{\Theta} | d_n, K)$, extracting the measured probability distribution
for a single parameter (say $\alpha$) involves integrating
\begin{equation}\label{eq:marginalize}
 p(\alpha | d_n, K) = \int\D \vec{\Theta}_\alpha\, p(\vec{\Theta} | d_n, K),
\end{equation}
where $\vec{\Theta}_\alpha$ is the set of remaining parameters, i.e.
$\vec{\Theta}_\alpha:=\vec{\Theta} - \{\alpha\}$.

@@

\section{How is PE affected if we ignore NS matter effects?}\label{s1:PEwithnoNS}
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}
\caption{We show here probability distributions for mass ratio $\eta$ as measured
for the same injection at different SNRs. The intrinsic parameters of the source
are: $q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and $\lambdans=2000$;
and the signal is injected at SNRs $\rho=\{20,30,50\}$ (left to right). The templates
{\it ignore} tidal effects.
In each panel: the dashed red line marks the median {\it measured} value
$\eta^\mathrm{Median}$, while the dashed green line show the true value
$\eta^\mathrm{Injected}$. The difference between the true and measured values
is the systematic bias introduced
by ignoring tidal effects in templates. The darker shading shows
the recovered $90\%$ confidence interval for $\eta$, $(\Delta\eta)^{90\%}$, 
which is a direct measure of our statistical uncertainty on $\eta$.
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
\label{fig:SingleSystemEtaPDFvsSNR}
\includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNMchirpCIWidths90_0_Lambda_SNR}\\
\includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNEtaCIWidths90_0_Lambda_SNR}\\
\includegraphics[trim = {2cm 0 0 0},width=2.1\columnwidth]{plots/TNChiBHCIWidths90_0_Lambda_SNR}
We show here the statistical uncertainty of our measurement of the 
dimensionless mass ratio $\eta$ (at $90\%$ confidence), over the NSBH
parameter space.
Individual panels show the same as a function of BH mass and spin.
Across each row, we see the effect of increasing signal strength (i.e.
SNR) with the NS's tidal deformability $\lambdans$ fixed. Down each
column, we see the effect of increasing $\lambdans$, at fixed SNR.
\label{fig:CIWidths90_Lambda_SNR}
\includegraphics[width=1.6\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\begin{figure*}[!t]
\includegraphics[width=2\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR20_70_linear}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\includegraphics[width=1.95\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
\label{fig:TN_BHspinBias_vs_Lambda_SNR}
Past and future GW search and source characterization efforts with Advanced LIGO
have used (or plan to) BH-BH waveform templates for NS-BH mergers, ignoring
NS tidal effects on the emitted GWs. In this section we present a fully
Bayesian analysis of the effects of this assumption on the recovery of NSBH
masses and spins from their GW signals.
